Clinical prediction with localized modeling using similarity-based cohorts: A scoping review

### 2023-09-09 docs/Manuscript/Manuscript.Rmd
make or note changes following email with adi

---
title: "Clinical prediction with localized modeling using similarity-based cohorts: A scoping review"
author:
  - name: Adi Cohen
    email: ac4569@my.nsu.edu
    affiliation: 1
  - name: Patti McCall-Junkin
    email: pattimccall.junkin@gmail.com
    affiliation: 2
  - name: Jason Cory Brunson
    email: jason.brunson@medicine.ufl.edu
    affiliation: 3
    correspondingauthor: true
address:
  - code: 1
    address: College of Osteopathic Medicine, Nova Southeastern University, 3200 S University Drive, Davie, FL 33328
  - code: 2
    address: Smathers Libraries Academic Research & Consulting Services, University at Florida, P. O. Box 117000, Gainesville, FL 32611-7000
  - code: 3
    address: Laboratory for Systems Medicine, University of Florida, P.O. Box 100225 JHMHC, Gainesville, FL 32610-0225
abstract: |
  Background:
  Applications of classical case-based reasoning (CBR) to clinical and health tasks have given rise to a family of techniques in which an (often conventional) model is fitted to a subset of labeled cases, selected for relevance or similarity to an unlabeled case to which the model is applied. These localized models, as we propose to call them, may perform as well as "black-box" models while retaining direct interpretations. Their adoption has been hindered by fragmented development and implementation challenges. Our goal in this scoping review was to describe the clinical and health applications of localized models to date and to derive a general framework for their design and evaluation.

@@

We expected the general framework to boil down to three choices: A patient similarity measure, a cohort selection process, and a statistical model family. Such a general method would be straightforward, if laborious, to implement, and specializations of it would account for the majority of real-world applications. A general implementation based on this method would allow researchers to expedite every step of the analysis process, including selection, optimization, and evaluation, and enable sensitivity, robustness, and multiverse analyses that help identify the most consequential choices along the way.

@@

Our methods followed prepared plans, which we describe in detail in the Appendix.
We also note deviations from these plans and the reasons for them.

@@

We include a PRISMA checklist as Supplemental Table 1 and a PRISMA abstract checklist as Supplemental Table 2.
Because we focus on methodologies rather than conceptual approaches or evidence, we deviate in some ways from PRISMA guidelines. In particular, those items of the checklist involving bias assessment and quantitative synthesis are intended for meta-analyses so did not apply to this study.
Because we are not aware of any standard procedures for conducting reviews and syntheses of methodology, no protocol was prepared for this study.

@@

## Full text review

@@

We do observe one clear pattern: All of the proposed methods that most evidently outperformed their comparators (Kohonen + CBR [@Wyns2004a], TWNFI [@Song2006c, @Kasabov2010], GSA [@Liang2015a], CBR + rules (+ DML) [@Nicolas2014a], Gaussian process regression [@Zhang2018a], JITL-ELM [@Ma2020a]) are hybrids of localized modeling (in some cases CBR) with other techniques, often metric learning. Though @CampilloGimenez2012a and [@Ng2015] report non-superior performance by such hybrids, the pattern suggests the importance of the similarity measure to the retrieval step.
Meanwhile, when proposed approaches targeted cohort demarcation or choice of predictive model (statistical CBR [@Park2006]; individualized logistic regression, decision tree, and random forest [@Lee2015; @Lee2017]), they did not consistently outperform comparators.

@@

**Write this section!**

1. Deduplication (Covidence)
    a. Specific to that source: title, creator, URL, DOI
    b. May overlap: journal, publication date, ISSN
2. Title
    a. Exclude non clinical/non medical setting
    b. Must be in English
    c. Must be original study (not reviews, surveys, opinion, news)
    d. Exclude if search term clearly has different meaning than intended
3. Abstract
4. Full Text
    a. Study criteria (inclusion/exclusion)

@@

Our selection of relevant papers from the search corpus will be based on the following inclusion/exclusion criteria:

@@

```{r tab:performance, echo=FALSE, message=FALSE, warning=FALSE}
# TODO: Place vertical separators between major columns.
tab_eval_comp |> 
  pivot_wider(
    id_cols = c(zotero_key, date_of_publication, measure, task),
    names_from = role, values_from = results
  ) |> 
  arrange(date_of_publication) |> 
  mutate(len = map2_int(proposal, comparator, ~ max(length(.x), length(.y)))) |> 
  mutate(across(c(proposal, comparator), ~ map2(., len, \(x, y) x[seq(y)]))) |> 
  unnest(c(proposal, comparator)) |> 
  separate(proposal, c("proposal", " "), sep = ": ") |> 
  separate(comparator, c("comparator", "  "), sep = ": ") |> 
  # use abbreviations throughout
  mutate(across(
    c(proposal, comparator),
    ~ str_replace(., "[Ll]ogistic [Rr]egression", "LR")
  )) |> 
  mutate(across(
    c(proposal, comparator),
    ~ ifelse(. == "Logistic", "LR", .)
  )) |> 
  mutate(across(
    c(proposal, comparator),
    ~ str_replace(., "[Rr]andom [Ff]orest", "RF")
  )) |> 
  mutate(across(
    c(proposal, comparator),
    ~ str_replace(., "[Kk][\\- ][Nn]earest [Nn]eighbor(s){0,1}", "KNN")
  )) |> 
  mutate(across(
    c(proposal, comparator),
    ~ str_replace(., "([Aa]rtificial ){0,1}[Nn]eural [Nn]et(work){0,1}", "NN")
  )) |> 
  mutate(across(
    c(proposal, comparator),
    ~ str_replace(., "[Oo]ne-class", "One-class")
  )) |> 
  mutate(across(
    c(proposal, ` `, comparator, `  `),
    ~ ifelse(is.na(.), "", .)
  )) |> 
  transmute(
    Citation = str_c("[@", zotero_key, "]"),
    Measure = measure, Problem = task,
    Proposals = proposal, ` `,
    # insert blank column to separate right- and left-justified entries
    # `   ` = "",
    Comparators = comparator, `  `
  ) |> 
  mutate(across(
    c(Citation, Measure, Problem),
    ~ ifelse(is.na(lag(.)) | . != lag(.), ., "")
  )) |>
  # prevent numeric list columns from dominating the horizontal space
  # mutate(across(c(Proposals, Comparators), ~ str_c("\\hspace{3em}", .))) |> 
  mutate(across(c(` `, `  `), ~ str_c(., "\\hspace{6em}"))) |> 
  knitr::kable(
    align = "llllrlr",
    caption = paste0(
      "\\label{tab:performance}",
      "Evaluations of proposed methods and comparisons to alternative methods.",
      " (See original studies for full names and descriptions.)"
    )
  )
```

>>>>

---
title: "Clinical prediction with localized modeling using similarity-based cohorts: A scoping review"
author:
  - name: Adi Cohen
    email: ac4569@my.nsu.edu
    affiliation: 1
  - name: Patti McCall-Junkin
    email: pattimccall.junkin@gmail.com
    affiliation: 2
  - name: Jason Cory Brunson
    email: jason.brunson@medicine.ufl.edu
    affiliation: 3
    correspondingauthor: true
address:
  - code: 1
    address: College of Osteopathic Medicine, Nova Southeastern University, 3200 S University Drive, Davie, FL 33328
  - code: 2
    address: Smathers Libraries Academic Research & Consulting Services, University at Florida, P. O. Box 117000, Gainesville, FL 32611-7000
  - code: 3
    address: Laboratory for Systems Medicine, University of Florida, P.O. Box 100225 JHMHC, Gainesville, FL 32610-0225
abstract: |
  Background:
  Applications of classical case-based reasoning (CBR) to clinical and health tasks have given rise to a family of techniques in which a statistical model is fitted to a subset of labeled cases, selected for relevance or similarity to an unlabeled case to which the model is applied. These localized models, as we propose to call them, may perform as well as "black-box" models while retaining direct interpretations. Their adoption has been hindered by fragmented development and implementation challenges. Our goal in this scoping review was to describe the clinical and health applications of localized models to date and to derive a general framework for their design and evaluation.

@@

We expected the general framework to come down to three choices: A patient similarity measure, a cohort selection process, and a statistical model family. Such a general method would be straightforward, if laborious, to implement, and specializations of it would account for the majority of real-world applications. A general implementation based on this method would allow researchers to expedite every step of the analysis process, including selection, optimization, and evaluation, and enable sensitivity, robustness, and multiverse analyses that help identify the most consequential choices along the way.

@@

Our methods followed prepared plans, which we describe in detail in the Appendix.
**Additional details are provided in Section \label{sec:appendix-selection}.**
We also note deviations from these plans and the reasons for them.

@@

**Correct "n = PubMed" in PRISMA diagram.**
We include a PRISMA checklist as Supplemental Table 1 and a PRISMA abstract checklist as Supplemental Table 2.
Because we focus on methodologies rather than conceptual approaches or evidence, we deviate in some ways from PRISMA guidelines. In particular, those items of the checklist involving bias assessment and quantitative synthesis are intended for meta-analyses so did not apply to this study.
Because we are not aware of any standard procedures for conducting reviews and syntheses of methodology, no protocol was prepared for this study.

@@

## Full text review
\label{sec:full-text}

@@

We do observe one clear pattern: All of the proposed methods that most evidently outperformed their comparators (Kohonen + CBR [@Wyns2004a], TWNFI [@Song2006c, @Kasabov2010], gravitational search algorithms (GSA) [@Liang2015a], CBR + rules (+ DML) [@Nicolas2014a], Gaussian process regression [@Zhang2018a], JITL-ELM [@Ma2020a]) are hybrids of localized modeling (in some cases CBR) with other techniques, often metric learning. Though @CampilloGimenez2012a and [@Ng2015] report non-superior performance by such hybrids, the pattern suggests the importance of the similarity measure to the retrieval step.
Meanwhile, when proposed approaches targeted cohort demarcation or choice of predictive model (statistical CBR [@Park2006]; individualized logistic regression, decision tree, and random forest [@Lee2015; @Lee2017]), they did not consistently outperform comparators.

@@

Most deduplication was done automatically in Covidence. As full-text review was done in Zotero, some additional duplicates were noticed and merged.

When abstracts were not obtained by search or by Covidence, we attempted to find them online using DOIs; when an abstract could not be found, screening was based on the title alone. We decided to screen titles and abstracts conservatively, rejecting only studies that were clearly outside the scope of our review. The reasons for rejection were four, as discussed in the main text:
    a. Exclude non clinical/non medical setting
    b. Must be in English
    c. Must be original study (not reviews, surveys, opinion, news)
    d. Exclude if search term clearly has different meaning than intended
(Two examples of (d) are the use of the term "personalized model" to refer in some cases to parameterized models tuned to individual patient measurements and in others to patient-centered models of care.) Studies that passed title/abstract screen were exported to Zotero.

@@

Some PDFs were obtained using Zotero from a university workstation, and the remaining were obtained through university library services. The review process is detailed in Section \label{sec:full-text}.

Our selection of relevant papers from the search corpus was based on the following inclusion/exclusion criteria:

@@

**TODO: Place vertical separators between major columns.**

```{r tab:performance, echo=FALSE, message=FALSE, warning=FALSE}
tab_eval_comp |> 
  pivot_wider(
    id_cols = c(zotero_key, date_of_publication, measure, task),
    names_from = role, values_from = results
  ) |> 
  arrange(date_of_publication) |> 
  mutate(len = map2_int(proposal, comparator, ~ max(length(.x), length(.y)))) |> 
  mutate(across(c(proposal, comparator), ~ map2(., len, \(x, y) x[seq(y)]))) |> 
  unnest(c(proposal, comparator)) |> 
  separate(proposal, c("proposal", " "), sep = ": ") |> 
  separate(comparator, c("comparator", "  "), sep = ": ") |> 
  # use abbreviations throughout
  mutate(across(
    c(proposal, comparator),
    ~ str_replace(., "[Ll]ogistic [Rr]egression", "LR")
  )) |> 
  mutate(across(
    c(proposal, comparator),
    ~ ifelse(. == "Logistic", "LR", .)
  )) |> 
  mutate(across(
    c(proposal, comparator),
    ~ str_replace(., "[Rr]andom [Ff]orest", "RF")
  )) |> 
  mutate(across(
    c(proposal, comparator),
    ~ str_replace(., "[Kk][\\- ][Nn]earest [Nn]eighbor(s){0,1}", "KNN")
  )) |> 
  mutate(across(
    c(proposal, comparator),
    ~ str_replace(., "([Aa]rtificial ){0,1}[Nn]eural [Nn]et(work){0,1}", "NN")
  )) |> 
  mutate(across(
    c(proposal, comparator),
    ~ str_replace(., "[Oo]ne-class", "One-class")
  )) |> 
  mutate(across(
    c(proposal, ` `, comparator, `  `),
    ~ ifelse(is.na(.), "", .)
  )) |> 
  transmute(
    Citation = str_c("[@", zotero_key, "]"),
    Measure = measure, Problem = task,
    Proposals = proposal, ` `,
    # insert blank column to separate right- and left-justified entries
    # `   ` = "",
    Comparators = comparator, `  `
  ) |> 
  mutate(across(
    c(Citation, Measure, Problem),
    ~ ifelse(is.na(lag(.)) | . != lag(.), ., "")
  )) |>
  # prevent numeric list columns from dominating the horizontal space
  # mutate(across(c(Proposals, Comparators), ~ str_c("\\hspace{3em}", .))) |> 
  mutate(across(c(` `, `  `), ~ str_c(., "\\hspace{6em}"))) |> 
  knitr::kable(
    align = "llllrlr",
    caption = paste0(
      "\\label{tab:performance}",
      "Evaluations of proposed methods and comparisons to alternative methods.",
      " (See original studies for full names and descriptions.)"
    )
  )
```
